Estimating the incidence of SARS-CoV-2 from prevalence data

### 2023-05-25 cis-perfect-testing.tex
Move text from sim study doc across

@@

Two major challenges arise when dealing with this data: double interval
censoring and left truncation. Double interval censoring arises because
we can only bound the start and end of an infection by observing an
individual's test results transitioning from negative to positive and
vice versa. Left truncation occurs because an individual is infected at
time $t$ is only observed if their infection duration is at least
$t^N - t$ days long, where $t^N$ is the time of their next test.

To address these challenges, we employ two distinct approaches: the
``individual'' model and the ``total'' model.

@@

This chapter proceeds as follows.
\begin{itemize}
\item  I introduce the motivating dataset in section~\ref{cpt:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
\item  In section~\ref{cpt:sec:model}, I develop a model accounting for double interval censoring and truncation. This builds on previous work in both biomedical and bird nesting studies. The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature. Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
\item  In section~\ref{cpt:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior. Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations. I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
\item  Section~\ref{cpt:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS. I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors. However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
\item  Section~\ref{cpt:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
\end{itemize}

\section{Data}\label{cpt:sec:data}

The CIS continually enrols individuals and tests them on an ongoing basis, following a
specified testing protocol. The protocol specifies that the initial five
tests are spaced every 7 days, then the spacing is every 28 days.
However, real-world considerations often lead to variations in test
times and missed tests. Our analysis focuses solely on the binary result
(positive or negative) of these tests, assuming no misclassification
bias

Episodes are taken from the data extract of the 3rd Jan 2023, as
computed by SARAH using her updated episode definition. Only tests taken
as part of CIS are included (ie: no pillar 2 testing). The following
criteria are applied.

\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test following the final positive in the episode has been
  recorded.
\end{itemize}

\section{Modelling the duration}\label{cpt:sec:model}

\begin{itemize}
\item
  We initially assume perfect testing (no misclassification), although
  later relax this assumption.
\item
  Time is discrete: $1, \dots, T$.
\item
  We observe $n_a$ episodes of infection, indexed by $i$.
\item
  The $i$ in which episode $i$ occurs is tested at times
  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
\item
  Each observed episode has an unknown start of infection time $b_i$
  and end of infection time $e_i$ (the first and last day that the
  individual would test positive due to this episode respectively).
  These are realisations of the random variables $B_i$ and $E_i$
  respectively.
\item
  The duration of the infection is the number of days for which an
  individual tests positive, $D_i = E_i - B_i + 1$.
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\item
  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
  result for individual $i$ at time $t$. Under the perfect testing
  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
\item
  Throughout, the convention that lower-case letters are realisations of
  upper-case random variables is used.
\end{itemize}

\subsection{Individual model}\label{cpt:sec:individual-model}

\subsubsection{Likelihood}\label{cpt:sec:likelihood}

@@

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (ie: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{cpt:fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{cpt:sec:deriv-probs}).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{cpt:fig:partitionSpace}}
\end{figure}

@@

\subsubsection{Full posterior}\label{cpt:sec:full-posterior}

@@

\subsection{Total model}\label{cpt:sec:total-model}

@@

\section{Parameterisation and priors for the survival function} \label{cpt:sec:parameters-priors}

\section{Simulation study} \label{cpt:sec:simulation-study}

\section{Conclusion} \label{cpt:sec:conclusion}

>>>>

\listoftodos

\pagebreak

@@

Two major challenges arise when dealing with this data: double interval
censoring and left truncation. Double interval censoring arises because
we can only bound the start and end of an infection by observing an
individual's test results transitioning from negative to positive and
vice versa. Left truncation occurs because an individual is infected at
time $t$ is only observed if their infection duration is at least
$t^N - t$ days long, where $t^N$ is the time of their next test.
\todo[inline]{Add figures demonstrating the censoring and truncation here}

To address these challenges, we employ two distinct approaches: the
``individual'' model and the ``total'' model.
\todo[inline]{\textbf{Should I cut one of the models?}}

\todo[inline]{Need to think about what of the below belongs here in the introduction, and turn these parts into a proper literature review}
\todo[inline]{\textbf{Is it better to have a large literature review here, or to discuss relevant literature in each section (the papers on survival analysis in the model section, the papers on priors in that section, COVID estimates where I use them, etc.)?}}

This chapter proceeds as follows.
\begin{itemize}
\item  I introduce the motivating dataset in section~\ref{perf-test:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
\item  In section~\ref{perf-test:sec:model}, I develop a model accounting for double interval censoring and truncation. This builds on previous work in both biomedical and bird nesting studies. The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature. Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
\item  In section~\ref{perf-test:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior. Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations. I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
\item  Section~\ref{perf-test:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS. I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors. However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
\item  Section~\ref{perf-test:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
\end{itemize}

\section{Problem description}\label{perf-test:sec:data}

Consider a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
We wish to estimate the distribution of time for which an individual is positive for.
An individual could have multiple infections, in which case we refer to "infection episodes", where each episode corresponds to a single infection event.

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
In this chapter I will assume that the time between infection and becoming detectable is negligible.
The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
An episode is \emph{detected} if they test positive at least once; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).

I take a survival analysis approach to analysing this data.
Survival analysis is the area of statistics which estimate the distribution of times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In this context, the distribution to estimate is that of duration of positivity.

The reason for this approach, over modelling biomarkers (as in chapter\todo{ref chapter on ATACCC}) is...

The methods in this chapter deal with modelling two issues that arise from the data. False negatives are not dealt with and left to chapter X.

The first issue is double interval censoring.
Explain issue, brief overview of literature.

The second issue is truncation.

These combination of issues are rarely encountered in biomedical data (ref).

I use the CIS data described in section\todo{ref description of CIS data}, with positive tests heuristically divided into episodes.
The episode definition was developed by Sarah Walker and is as follows~\citePersonalComms:
\todo[inline]{Insert episode definition}

I filter the episodes as follows.
\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test to end the episode (ie: following the final positive in the episode) has been recorded.
\end{itemize}

\todo[inline]{Descriptive analyses of the data here}

\section{Modelling the duration}\label{perf-test:sec:model}

@@

\subsection{Notation and assumptions}
\todo[inline]{Some of this will possibly be defined elsewhere, such as in the introduction where CIS is introduced}
\begin{itemize}
\item
  Time is discrete: $1, \dots, T$.
\item
  We observe $n_a$ episodes of infection, indexed by $i$.
\item
  The $i$ in which episode $i$ occurs is tested at times
  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
\item
  Each observed episode has an unknown start of infection time $b_i$
  and end of infection time $e_i$ (the first and last day that the
  individual would test positive due to this episode respectively).
  These are realisations of the random variables $B_i$ and $E_i$
  respectively.
\item
  The duration of the infection is the number of days for which an
  individual tests positive, $D_i = E_i - B_i + 1$.
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\item
  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
  result for individual $i$ at time $t$. Under the perfect testing
  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
\item
  Throughout, the convention that lower-case letters are realisations of
  upper-case random variables is used.
\end{itemize}

\subsection{Individual model}\label{perf-test:sec:individual-model}

\subsubsection{Likelihood}\label{perf-test:sec:likelihood}

@@

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (ie: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{perf-test:sec:likelihood}).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{perf-test:fig:partitionSpace} \todo[inline]{Digitalise figure of regions.}}
\end{figure}

@@

\subsubsection{Full posterior}\label{perf-test:sec:full-posterior}

@@

\subsection{Total model}\label{perf-test:sec:total-model}

@@

\section{Parameterisation and priors for the survival function} \label{perf-test:sec:parameters-priors}

\todo[inline]{Section is currently notes: expand into prose which can largely come from previous documents}

Define hazard and survival (maybe this should be done previously?).

The hazard is convenient to use because it is unconstrained except in the interval [0 ,1]; include REFs.
Survival needs to be monotonic.

Discuss literature on Beta Process priors (maybe this should be in a broader literature review part?).

\subsection{Independent priors}

Traditional uninformative priors, such as Beta(1, 1) or Beta(0.5, 0.5), are unsuitable.

Instead propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.

\subsection{Smoothing priors}

\subsection{Informative priors}

ATACCC can be used to provide prior information for the CIS analysis.

Intuition: ATACCC informs early part, CIS informs late part.

Problem: do not believe late part of ATACCC analysis.

First, transform ATACCC analsysis into a posterior over survival time.

Second, approximate ATACCC analysis with multivariate normal distribution.
Show that this is a good approximation.

Finally, use a Beta Process to "discount" some of the information in ATACCC towards a vague prior (as in the previous section).

\section{Simulation study} \label{perf-test:sec:simulation-study}

\subsection{Setup}

\subsubsection{Testing and infection times}

\subsubsection{Duration of positivity}

We require a ground truth for the distribution giving duration of positivity.
The majority of this distribution is based on \autoref{ATACCC}'s estimates from the ATACCC data.
However, the tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
These curves and the combined curve are compared in figure \@ref(fig:duration-curves).
Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.

Sarah uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function (@roystonSTPM; @roystonFlexible) with covariates introduced via proportional odds.

We know of two biases introduced from this analysis: “There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test. Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short). However, there is another bias due to the survey missing positive episodes entirely if they are short. This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes. This will bias the clearance curves upwards (making the estimates too long).”~\autocite{cisMethodsONS}.

To form the duration distribution used in the simulation, we combine these two estimates.
The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
Define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t-3) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.

\todo[inline]{Figure with duration distributions}

\subsubsection{Test results and episode summaries}
\todo[inline]{Make notation here consistent with elsewhere}

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside of this period, all tests are negative (ie: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).

Each individual is then summarised into an "episode" (when working with CIS data, the same individual can have multiple episodes due to reinfections, these are split heuristically but a negligible number in the pre-Omicron period and hence ignored within the simulation).
An episode is characterised by:

- The total number of positive tests observed per individual, $n_i^\text{pos}$.
- The time of an individual's first positive test $V'_i$.
- The time of an individual's last negative test prior to $V_i'$, $E'_i$.
- The time of an individual's last positive test, $L_i'$.
- The time of an individual's first negative test after $L_i'$, $R_i'$.

\subsubsection{Restrictions and sample size}

When using the real testing schedules, we then filter the episodes to replicate the process used when working with the CIS data.
First, we discard any episodes where $n_i^\text{pos}=0$ as they are never detected.
We then apply the same criteria as that applied to the modelling dataset (see Section \@ref(design)).
Finally, of those remaining episodes, we select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
The last step is required because we infect everyone in the simulation which does not happen in reality.

In the fixed testing schedules, there is no automatic choice of sample size.
We simulate a large number of episodes, and then choose to keep 4800 episodes\todo{Check how many this is} where at least one positive test was returned.

\subsection{Descriptive analyses of simulated data}

\subsection{Results}

\todo[inline]{Need to decide what exactly to include... There's lots of possible results but need to form some coherent story here!}

\section{Conclusion} \label{perf-test:sec:conclusion}

### 2023-05-25 main.tex
Format TeX and working pipeline
Sort chapter structure
Start drawing together writings into appropriate sections
Move text from sim study doc across

\documentclass{article}
\usepackage{graphicx}
\graphicspath{{images/}}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{textcomp} % provide euro and other symbols

@@

\usepackage{subfiles} % Best loaded last in the preamble

\title{Estimating the incidence of SARS-CoV-2 from prevalence data}
\author{Joshua Blake}
\date{2023}

@@

\maketitle
abc
\section{Chapter 1}

\subfile{chapter1/chapter1}

\begin{align}
    \cov
\end{align}

>>>>

\documentclass[a4paper]{report}
\usepackage{graphicx}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{textcomp} % provide symbols
\usepackage{todonotes}
% \usepackage{microtype}

@@

\newcommand{\citePersonalComms}{(personal communication)}

\title{Estimating the incidence of SARS-CoV-2 from prevalence data}
\author{Joshua Blake}
\date{\today}

\usepackage[backend=biber,style=authoryear,uniquename=false]{biblatex}

\usepackage{subfiles} % Best loaded last in the preamble
\addbibresource{references.bib}

@@

\maketitle

\chapter{Text to use}

\section{Data}
The CIS continually enrols individuals and tests them on an ongoing basis, following a
specified testing protocol. The protocol specifies that the initial five
tests are spaced every 7 days, then the spacing is every 28 days.
However, real-world considerations often lead to variations in test
times and missed tests. Our analysis focuses solely on the binary result
(positive or negative) of these tests, assuming no misclassification
bias.

The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
A full description of the study can be found in the supplementary materials of [Pouwels et al. 2020](https://www.thelancet.com/journals/lanpub/article/PIIS2468-2667(20)30282-6/fulltext#supplementaryMaterial) or [the study protocol](https://www.ndm.ox.ac.uk/covid-19/covid-19-infection-survey/protocol-and-information-sheets).
In brief, households are invited to the study from databases held by the ONS.
The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
Within a household, all indvidiuals aged 2 and over were invited to participate.
Once invited, an enrolement swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolement) after which monthly swabs are taken.

During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
See figure \@ref(fig:test-schedules) for further detail.

\todo[inline]{Add some descriptive figures eg those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}

\section{Producing duration from ATACCC}

The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated ([see this Overleaf paper for further details](https://www.overleaf.com/9549594748zbvpgjgvmgsd)).
By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
calculated for the posterior sample $i$ as:
$$
f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
$$
Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
I have used $N = 1000$ and $n = 100,000$, 
Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
However, the results here do not use this for consistency.

\chapter{Survival analysis with perfect testing} \label{perf-test}

\subfile{cis-perfect-testing}

\chapter{Survival analysis with misclassification} \label{imperf-test}

\subfile{cis-imperfect-testing}
